---
title: "Intro to R day 6"
output:
  pdf_document: default
  html_notebook: default
---


We'll do this day in R Notebook. 

To create a new chunk either choose Insert - R from the menu right above this pane, or press cntrl alt i from the keyboard.

And you can now write regular R code into the chunk.
```{r message=FALSE}
library(tidyverse)
piaac <- read.csv("http://www.ut.ee/~iseppo/piaacest.csv")
summary(piaac$gender)
```


#Lists
We will continue with another data type in R, called lists:

Lets create a nr of R objects and then create a list from them:
```{r}
name <- "Masha"
surname <- "Mishka"
length <- 1.65
peripherals <- data.frame(nr.of.fingers =c(5, 5),
                          nr.of.toes=c(5,5))
peripherals

mashadata <- list(first.name=name, 
                  surname=surname, 
                  length=length,
                  peripherals=peripherals)
mashadata
``` 

Why do we need lists? A lot of R functions give you lists as a result. 

I am showing this with some regular regression:

```{r}
piaac <- read.csv("http://www.ut.ee/~iseppo/piaacest.csv")
```


```{r}
regr1 <- lm(data=piaac, log(earnmth) ~ gender + pvnum1)
regr1
summary(regr1)
```

But this regr1 object is actually a list, containing a lot of additional information:

```{r}
str(regr1)
```

```{r}
effectMale <- regr1[["coefficients"]]["genderMale"]
```

You can now use this inline aswell: Effect for males is `r round(effectMale,2)`.

So -- to access data objects put into a list, you will need to use double square brackets: `listname[[objectname]]`

```{r}
mashadata[["first.name"]]
```

Lets try the same with t-test:

```{r}
t.test(piaac$pvlit1)
```

It is actually a list, a list from where you can get the evaluated parameters out of:
```{r}
result <- t.test(piaac$pvlit1)
str(result)
result[["conf.int"]]
```

```{r}
lower <- result[1]
upper <- t.test(piaac$pvlit1)[["conf.int"]][2]
class(result)
```


#Functions

Creating functions in R is actually really easy:

```{r}
multiply_by_3 <- function(x){
  result <- x*3
  return(result)
}
multiply_by_3(5)
```

```{r}
multiply_by_3(4)
```

```{r}
y<-c(3,4,5)
multiply_by_3(y)
```

```{r}
multiplydivide <-function(x, multiplier =3, divisor =1){
  result <- x * multiplier/ divisor
  return(result)
}
```

```{r}
multiplydivide(x =4)
```


```{r}
multiplydivide( x = 4, multiplier = 1, divisor = 2)
```


```{r}
multiplydivide
```

Just remember to load the function in, once you have written it. And you can start using it.




Your turn:
```{r}

substract <- function(a, b){
  result <- a-b
  return(result)
}

substract(5, 8)
```

```{r}

findlower <- function(x){
  ttestresult <- t.test(x)
  lower <- ttestresult[["conf.int"]][1]
  return(lower)
}

findupper <- function(x){
  ttestresult <- t.test(x)
  lower <- ttestresult[["conf.int"]][2]
  return(lower)
}
```

```{r}
findupperorlower <- function(x, whichone="lower"){
  ttestresult <- t.test(x)
  if(whichone=="lower"){
  answer <- ttestresult[["conf.int"]][1]
  }
  if(whicone=="upper"){
  answer <- ttestresult[["conf.int"]][2]
  }
  return(answer)
}
```

```{r}
library(tidyverse)
piaac %>%
  group_by(gender)%>%
  summarize(lowernum=findlower(pvnum1), avergenum=mean(pvnum1, na.rm=T), uppernum=findupper(pvnum1))
  
```


You can save your functions into an R script file and then just source the file to have them available:

```{r, eval=FALSE}
source("functions2.R")
```

source() runs the file and performs everything that is written into it. 

# Tidy data, data in wide and long format

R loves to have data in tidy format, one measurement per line.

```{r}
library(eurostat)
gdpdata <- get_eurostat("tec00114")
gdpdata
```


This is data in wide format:
```{r}

animals <- read.csv("http://www.ut.ee/~iseppo/animals.csv")
animals
```

To convert it to long format, you would use gather() from package tidyr:

```{r}
gather(animals, key="measurement", value="value",
       length, width,age)
```

```{r}
gdp <- read.csv("http://www.ut.ee/~iseppo/gdpestimate.csv")
gdp
library(lubridate)
gdp$date <- ymd(gdp$date)
```

```{r}
ggplot(gdp, aes(x=date))+
  geom_line(aes(y=firstEstimate))+
  geom_line(aes(y=latestEstimate), color="red")
```

```{r}
gdp.long <- gather(gdp, key=measurement, value=value, firstEstimate,
                   latestEstimate)
head(gdp.long)
ggplot(gdp.long, aes(x=date))+
  geom_line(aes(y=value, color=measurement))
```

Spread
```{r}
gdpdata <- get_eurostat("tec00114")
gdpdata
gdpdata_wide <- spread(gdpdata, key=geo, value=values)
library(knitr) # for function "kable"
kable(head(gdpdata_wide))
```

# Joining data

```{r}
animallength <- data.frame(animal=c("cat", "dog", "elephant"),
                           length=c(10,20,50))
animalwidth <- data.frame(animal=c("cat", "dog", "bear"),
                          width=c(5, 15, 10))

animallength <- read.csv("http://www.ut.ee/~iseppo/animallength.csv")
animalwith <- read.csv("http://www.ut.ee/~iseppo/animalwidth.csv")
animallength
```

```{r}
animalwidth
```

```{r}
anti_join(animalwidth, animallength)
```

```{r}
left_join(animalwidth, animallength,
          by=c("animal"="animal"))
```


Your turn at page 16:
```{r}
library(readr)
gdppercap <- read_csv("http://www.ut.ee/~iseppo/gdppercap.csv")

population <- read_csv("http://www.ut.ee/~iseppo/population.csv")
```

lets take a look:

```{r}
summary(gdppercap)
summary(population)
```

```{r}
countries <- full_join(gdppercap, population)
```

```{r}
anti_join(population, gdppercap)
```

```{r}
q1.2017 <- read_csv("http://www.ut.ee/~iseppo/q12017.gz")
q2.2017 <- read_csv("http://www.ut.ee/~iseppo/q22017.gz")
q3.2017 <- read_csv("http://www.ut.ee/~iseppo/q32017.gz")
head(q1.2017)

tax.2017 <- bind_rows(q1.2017, q2.2017, q3.2017)
dim(tax.2017)
```

```{r}
county_employees <- tax.2017%>%
  group_by(county, quarter)%>%
  summarize(nrOfEmployees=sum(employees, na.rm=T))
county_employees

county_employees_wide <- spread(county_employees, key=quarter, value=nrOfEmployees)
county_employees_wide
```

```{r}
county_employees_wide %>%
  mutate(change = q3.2017-q1.2017)%>%
  arrange(change)
```



